Empirical Bayes analysis of \RNAseq{} data for detection of gene expression heterosis

### 2015-07-14 ebayes_rnaseq_heterosis.tex
update volcano plot and name
up to discussion

\begin{abstract}
Heterosis, or hybrid vigor, refers to the enhancements in the phenotype of hybrid progeny relative to their inbred parents. Although the heterosis phenomenon is extensively utilized in agriculture, the molecular basis of heterosis is still largely unknown. In an effort to understand phenotypic heterosis at the molecular level, researchers have begun to measure transcript abundance levels of thousands of genes in parental inbred lines and their hybrid offspring using RNA sequencing (\RNAseq{}) technology.  The resulting data allow researchers to search for evidence of gene expression heterosis as one potential molecular mechanism underlying heterosis of agriculturally important traits.  The null hypotheses of greatest interest in testing for gene expression heterosis are composite null hypotheses that are difficult to test with standard statistical approaches for \RNAseq{} analysis. To address these shortcomings, we develop a hierarchical model based on a negativeb binomial data model and draw inferences using a computationally tractable empirical Bayes approach to inference. We demonstrate improvements over alternative methods via a simulation study based on a maize experiment and then analyze that maize experiment with our newly proposed methodology. This article has supplementary material online.
\end{abstract}

@@

Recently, \cite{ji2014estimation} introduced an approach to assess gene expression heterosis using microarray data under the assumption that these data are continuous. They built a normal hierarchical model for microarray measurements of transcript abundance that allows borrowing of information across genes to estimate means and variances. They introduced an empirical Bayes framework that first estimates model hyperparameters, then estimates the posterior distribution for gene-specific parameters conditional on those hyperparameters, and finally computes heterosis probabilities based on integrals of regions under this posterior. This development was necessary due to the composite null hypotheses in tests for heterosis. These hypotheses, which many available methods do not fully accommodate, remain a challenge in the transition from continuous measurements of transcript abundance to count-based measurements that arise from RNA sequencing (\RNAseq{}) technology. Building on the work of \citeauthor{ji2014estimation} with the normal data model, we construct a hierarchical model based on a negative-binomial data model. We also utilize an empirical Bayes approach to obtain \emph{maximum a posterior} (MAP) estimates of the hyperparameters and the posterior distributions for the gene-specific parameters conditional on those hyperparameters.

@@

We separate the parameters of the model in Section \ref{s:model} into gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ where $\theta_g = (\phi_g,\alpha_g,\delta_g,\psi_g)$, normalization factors $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$, and hyperparameters $\pi = (\eta, \sigma)$ where $\eta=(\eta_\phi,\eta_\alpha,\eta_\delta)$ and $\sigma=(\sigma_\phi,\sigma_\alpha,\sigma_\delta,\sigma_\psi)$. We obtain estimates for the hyperparameters and then base gene-specific inference on the posterior conditional on these estimates.

To obtain normalization factors $\hat{\gamma}$, we use the weighted trimmed mean of $M$ values (TMM) method of \cite{robinson2010scaling}. We then use the methods of \cite{mccarthy2012differential}, to obtain genewise dispersion estimates, $\hat{\phi}_g$, and their generalized linear model methods to obtain estimates for the remaining gene-specific parameters, i.e. $\hat{\phi}_g$, $\hat{\alpha}_g$, and $\hat{\delta}_g$. Using the gene-specific parameter estimates, we estimate hyperparameters for the location and scale parameters in the hierarchical model using a central method of moments approach, e.g. $\hat{\eta}_\delta = \sum_{g=1}^G \hat{\delta}_g/G$ and $2\hat{\sigma}_\delta^2 = \sum_{g=1}^G (\hat{\delta}_g - \hat{\eta}_\delta)^2/(G-1)$.

@@

To perform this Bayesian analysis, we used the statistical software \Stan which implements a variant of Markov chain Monte Carlo (MCMC) called Hamiltonion Monte Carlo (HMC) \citep{neal2011mcmc} to obtaining samples from the posterior in equation \eqref{e:condind}. We ran 4 simultaneous chains with random initial starting values for 1000 burnin (and tuning) iterations followed by another 1000 iterations thinned by 4 (to reduce storage space) for inference. We monitored convergence using the potential scale reduction factor (PSRF) \citep{Gelm:Rubi:infe:1992} for $\phi_g$, $\alpha_g$, $\delta_g$, and $\psi_g$. If the maximum PSRF was greater than 1.1, we reran the chain with an additional 500 iterations (250 burnin and 250 for inference) and repeated this process until all PSRF's were less than 1.1.

@@

\subsection{Implementation in \ShrinkBayes{}}

@@

We applied the \edgeR{} package \citep{robinson2010edgeR} from Bioconductor \citep{gentleman2004bioconductor} to the maize data on 27,888 genes with average count at least one and at most two zero read counts for each variety across four replications.  This analysis produced sample-specific normalization factors using the TMM method \citep{robinson2010scaling},
%(0.080, -0.041, -0.054, 0.000, -0.036, -0.147, 0.093, 0.059, -0.004, -0.057, 0.067,  0.041) 
gene-specific dispersion estimates, and gene-variety-specific location parameter estimates \citep{mccarthy2012differential}. We treated these estimates as the true parameter values for the simulation study so that our simulated datasets mimicked the existing covariance structure among the gene-variety means of the original maize data.

Using these parameters and normalization factors, we simulated data according to the negative-binomial model in \eqref{e:data} independently for each gene. For each simulation, we analyzed a subset of 25,000 genes selected randomly from genes with simulated counts at least one on average and with at most two zeros for each variety across replications.  We repeated this simulation process 10 times for each of 4, 8, and 16 replicates per variety, reusing normalization factors when necessary.  
% From Dan "Did we consider changing the number of zeros allowed when the number of reps is higher?  I like the idea of simplifying all this by just picking an average count threshold that his high enough to avoid problems.  I think that may be what others have done in the past."

@@

\edgeR{} can be used to test for differential expression between any two varieties based on the fit of a negative binomial log-linear model \citep{robinson2007moderated, robinson2010edgeR}. To construct a measure of heterosis, we computed the maximum likelihood estimates of the $\mu_{gv}$ parameters for all genes using \edgeR{}'s built-in Fisher scoring algorithm, and then used likelihood ratio tests to calculate two $p$-values for each gene: $p_{g1}$  for testing $H_{g01}:\mu_{g1}=\mu_{g3}$ and $p_{g2}$ for testing $H_{g02}:\mu_{g2}=\mu_{g3}$. Then, for each gene, we define a new $p$-value denoted as $p_g$ and determined as $p_g=1$ if the estimate of $\mu_{g3}$ falls between the estimates of $\mu_{g1}$ and $\mu_{g2}$ and $p_g=\max\{p_{g1},p_{g2}\}/2$ otherwise.  (For all relevant signifcance thresholds $\omega$ near 0, it can be shown that rejecting the null hypothesis of no heterosis of expression for gene $g$ whenever $p_g\leq \omega$ results in a test that is asymptotically size $\omega$.)  We then use $1-p_g$ as a final measure positively associated with strength of evidence for gene expression heterosis.

{\tt baySeq} allows for a wider range of hypotheses for each gene, including $H^*_{g1}: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H^*_{g 2}: \mu_{g1}=\mu_{g2}$, $H^*_{g 3}: \mu_{g1}=\mu_{g3}$, $H^*_{g 4}: \mu_{g2}=\mu_{g3}$, and $H^*_{g 5}:$ all $\mu_g$'s are distinct. In a technique similar to our application of \edgeR{}, we used the posterior probabilities of these hypotheses to construct a measure of heterosis for each gene. We set this measure to zero if the maximum likelihood estimate, calculated using \edgeR{}, of $\mu_{g3}$ is between the maximum likelihood estimates of $\mu_{g1}$ and $\mu_{g2}$. Otherwise, the measure is the sum of the posterior probabilities of $H^*_{g 2}$ and $H^*_{g 5}$, the only hypotheses that allow for heterosis.

@@

For each of the methods above, as well as the method described in Section \ref{s:gene_heterosis}, we sorted genes according to the computed measure of the strength of evidence for heterosis. From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to distinguish genes with heterosis, as defined in Section \ref{s:sim_data}, 
% From Dan, "why this citation?"
from those without heterosis \citep{landau2013dispersion}. A representative set of ROC curves is shown in Figure \ref{f:roc}. 
\begin{figure}[htbp]
\centerline{\includegraphics[width=\textwidth]{exampleROC0_1}}
\caption{Representative ROC curves for false positive rates below 0.1 for the approaches using \edgeR{}, {\tt baySeq},  and \ShrinkBayes{} described in Section \ref{s:alternative}, along with the independence and dependence methods using \RStan{} described in Section \ref{s:gene_heterosis}.}
\label{f:roc}
\end{figure}
For this simulation, we can see that the approaches based on the model in Section \ref{s:model}, i.e. \RStan{} and \ShrinkBayes{}, provide the best true positive rate for a given false positive rate. Also, as expected, as the sample size increases, our ability to distinguish genes with heterosis from genes without improves.

Figure \ref{f:auc} provides the area under the ROC curve below a false positive rate of 0.1 across the 10 simulations for each of the 3 different sample sizes. 
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{auc-facet-TRUE}}
\caption{Area under the ROC curves below a false positive rate of 0.1 for 3 different sample sizes for the approaches using \edgeR{}, {\tt baySeq},  and \ShrinkBayes{} described in Section \ref{s:alternative} , along with the independence and dependence methods using \RStan{} described in Section \ref{s:gene_heterosis}. Each line is a different simulation while the blue box indicates mean AUC and mean AUC plus and minus one standard error across the simulations.}
\label{f:auc}
\end{figure}
The pattern appears to hold, with the \RStan{} and \ShrinkBayes{} techniques outperforming the other two. %While there does not appear to be a significant difference between \ShrinkBayes{} and the independence model using \RStan{}, performance for \RStan{} appears to decrease for the dependence model where an unstructured covariance is estimated, at least for smaller sample sizes.

@@

We used our method to analyze a maize data set \citep{paschold2012complementation} on gene expression in parental lines B73 and Mo17 and the hybrid genotype (B73$\times$Mo17). The data are available in the Supplemental Materials. Each variety had four biological replicates measured with Illumina methodology and equipment. Reads were mapped to the whole reference genome using the short reads aligner, NOVOALIGN. For more specifics, please see \cite{paschold2012complementation}.

As explained in Section~\ref{s:sim_data}, we analyzed a set of $G=27,888$ genes for $V=3$ varieties and $n_v=4$ replicates ($v=1,2,3$).

We found MAP estimates for both the independence and dependence models as shown in Table \ref{t:map}. 
\input{include/map-laplace}
The MAP estimates for the hierarchical means, $\eta$, and standard deviations, $\sigma$, are consistent between the independence and dependence models with, perhaps, a slightly larger mean and smaller standard deviation for the logarithm of the overdispersion parameter, $\psi_g$, in the dependence model relative to the independence model. In the dependence model, there appears to be high correlation amongst $\mu_{g1}$, $\mu_{g2}$, and $\mu_{g3}$, and negative correlation between $\psi_g$ and each of $\mu_{g1}$, $\mu_{g2}$, and $\mu_{g3}$. The normalization factors, and their variance, are consistent between the two models, with the estimates for the dependence model being consistently 0.02 larger than the estimates for the independence model, and both indicating that there is about a 25\% increase in counts between the least and most thoroughly sequenced samples.

Independently for each gene, we ran four Markov chains conditional on the estimated hyperparameters to obtain approximate posterior distributions for all gene-specific parameters. Each Markov chain used initial values randomly drawn by {\tt Stan} and was run for a total of 2000 iterations, with 1000 iterations of burn-in and the remaining 1000 draws thinned by a factor of 4 for storage reasons. 
% From Dan "Do we need to be more specific about what method was used to generate the chains?"
The potential scale reduction factor, $\hat{R}$, was computed for each parameter \citep{Gelm:Rubi:infe:1992, Broo:Gelm:gene:1997}. If the maximum $\hat{R}$ was larger than 1.1 for any parameter, the total number of iterations was increased by 500, with the burn-in increased proportionally, and the chain was rerun. This process was repeated until the maximum $\hat{R}$ was less than 1.1 for all parameters.

With posteriors for all parameters, we can calculate empirical Bayes posterior probabilities for LPH and HPH. For each gene, the quantity of interest is likely to be the maximum of these two probabilities. For each gene with a high probability of either HPH or LPH, the magnitude of the effect is of interest. We calculate 
\begin{equation}
% \mbox{estimated effect size}_g = \left\{ 
% \begin{array}{ll}
% \hat{\mu}_{g3} - \min(\hat{\mu}_{g1},\hat{\mu}_{g2}) & \mbox{if } \hat{\mu}_{g3} < \min(\hat{\mu}_{g1},\hat{\mu}_{g2}) \\
% \hat{\mu}_{g3} - \max(\hat{\mu}_{g1},\hat{\mu}_{g2}) & \mbox{if } \hat{\mu}_{g3} > \max(\hat{\mu}_{g1},\hat{\mu}_{g2}) \\
% 0 & \mbox{otherwise},
% \end{array} 
% \right. 
\mbox{estimated effect size}_g \equiv \left\{ 
\begin{array}{ll}
\hat{\delta}_g - |\hat{\alpha}_g| & \mbox{if } \hat{\delta}_g > \phantom{-}|\alpha_g| \\
\hat{\delta}_g + |\hat{\alpha}_g| & \mbox{if } \hat{\delta}_g < -|\alpha_g| \\
0 & \mbox{otherwise},
\end{array} 
\right. 
\label{e:effect_size}
\end{equation}
where $\hat{\mu}_{gi}$ is the estimated posterior mean of $\mu_{gv}$ for $v=1,2,3$. Thus, the estimated effect size is the difference between hybrid mean and the nearest parent with negative values indicating LPH and positive values indicating HPH. If the hybrid mean is estimated to be between the parents, this effect is defined to be zero.

Figure \ref{f:volcano} provides a volcano plot, in this case a bivariate histogram, to visualize the maximum of the probabilities of LPH and HPH versus estimated effect size. The figure shows a peak at zero for probabilities below 0.5 with estimated effect sizes of zero. Above a probability of 0.5, we see a prototypical volcano pattern, with increased heterosis probability corresponding to larger estimated effect sizes and no estimated effect sizes near zero for genes with high heterosis probability. We also see asymmetry, with larger negative effect sizes than positive effect sizes due to genes with hybrid counts near zero and relatively high parental counts. Genes with high estimated posterior probabilities of heterosis and large estimated effect sizes are candidates for further investigation.

\begin{figure}
\centerline{\includegraphics[width=\textwidth]{maize_volcano_dexp}}
\caption{A bivariaet histogram of the maximum of the LPH and HPH probabilities versus estimated effect size defined in equation \ref{e:effect_size} for the B73 $\times$ Mo17 maize experiment.}
\label{f:volcano}
\end{figure}

With almost 30,000 genes in our data set, we can evaluate the distribution assumptions for the gene-specific parameters. Figure \ref{f:scatterplot} provides marginal and bivariate histograms for the gene-specific parameters.

\begin{figure}
\centerline{\includegraphics[width=\textwidth]{scatterplot_dexp}}
\caption{Marginal and bivariate histograms for gene-specific parameter posterior meansfor the B73 $\times$ Mo17 maize experiment. For reference, the marginal histograms have a dashed line indicating the appropriate marginal probability density functions in the hierarchical model with the MAP estimates.}
\label{f:scatterplot}
\end{figure}

Marginal and bivariate histograms for the model with an estimated covariance look similar although the estimated distribution $N(\hat{\eta},\hat{\Sigma})$ has increased variance for $\mu$ and decreased variance for $\psi$ as was previously seen in Table \ref{t:map} (results not shown).

@@

Geneticists speculate that gene expression heterosis is one possible explanation of phenotypic heterosis of traits, such as plant height or grain yield. Existing methods for identifying differential gene expression based on \RNAseq{} data are not directly applicable for detecting heterosis genes. \cite{ji2014estimation} introduced an empirical Bayes approach based on a hierarchical model for microarray data. We followed their approach, modified to allow for \RNAseq{} read counts as measures of transcript abundance. We developed an empirical Bayes approach based on obtaining MAP estimates for hyperparameters followed by MCMC to estimate gene-specific parameters. This methodology can be conveniently implemented in the statistical software, {\tt Stan}. The empirical Bayes posteriors can be used to estimate posterior probabilities of high and low parent heterosis.

Through a simulation study, we demonstrated that this method outperformed alternative methods (that are not designed to detect heterosis), and performed comparably well with an implementation of our model in \ShrinkBayes{}, which estimates the full posterior via INLA. We then demonstrated the use of the methodology on a maize experiment in which phenotypic heterosis is well known. 
% Estimates of heterosis to compare with Ji et. al.
In addition to providing estimates of heterosis probabilities and effect sizes, this real data analysis exposed violations of our normality assumptions, indicating room to improve the model and increase the power to detect gene expression heterosis. Some of our ongoing and future research will develop the details of such an extension.

>>>>

\begin{abstract}
Heterosis, or hybrid vigor, refers to the enhancements in the phenotype of hybrid progeny relative to their inbred parents. Although the heterosis phenomenon is extensively utilized in agriculture, the molecular basis of heterosis is still largely unknown. In an effort to understand phenotypic heterosis at the molecular level, researchers have begun to measure transcript abundance levels of thousands of genes in parental inbred lines and their hybrid offspring using RNA sequencing (\RNAseq{}) technology.  The resulting data allow researchers to search for evidence of gene expression heterosis as one potential molecular mechanism underlying heterosis of agriculturally important traits.  The null hypotheses of greatest interest in testing for gene expression heterosis are composite null hypotheses that are difficult to test with standard statistical approaches for \RNAseq{} analysis. To address these shortcomings, we develop a hierarchical negative binomial model and draw inferences using a computationally tractable empirical Bayes approach to inference. We demonstrate improvements over alternative methods via a simulation study based on a maize experiment and then analyze that maize experiment with our newly proposed methodology. This article has supplementary material online.
\end{abstract}

@@

Recently, \cite{ji2014estimation} introduced an approach to assess gene expression heterosis using microarray data under the assumption that these data are continuous. They built a normal hierarchical model for microarray measurements of transcript abundance that allows borrowing of information across genes to estimate means and variances. They introduced an empirical Bayes framework that first estimates model hyperparameters, then estimates the posterior distribution for gene-specific parameters conditional on those hyperparameters, and finally computes heterosis probabilities based on integrals of regions under this posterior. This development was necessary due to the composite null hypotheses in tests for heterosis. These hypotheses, which many available methods do not fully accommodate, remain a challenge in the transition from continuous measurements of transcript abundance to count-based measurements that arise from RNA sequencing (\RNAseq{}) technology. Building on the work of \citeauthor{ji2014estimation} with the normal data model, we construct a hierarchical model based on a negative binomial data model. We also utilize an empirical Bayes approach to obtain \emph{maximum a posterior} (MAP) estimates of the hyperparameters and the posterior distributions for the gene-specific parameters conditional on those hyperparameters.

@@

We separate the parameters of the model in Section \ref{s:model} into gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ where $\theta_g = (\phi_g,\alpha_g,\delta_g,\psi_g)$, normalization factors $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$, and hyperparameters $\pi = (\eta, \sigma)$ where $\eta=(\eta_\phi,\eta_\alpha,\eta_\delta, \eta_\psi)$ and $\sigma=(\sigma_\phi,\sigma_\alpha,\sigma_\delta,\sigma_\psi)$. We obtain estimates for the hyperparameters and then base gene-specific inference on the posterior conditional on these estimates.

To obtain normalization factors $\hat{\gamma}$, we use the weighted trimmed mean of $M$ values (TMM) method of \cite{robinson2010scaling}. We then use the methods of \cite{mccarthy2012differential}, to obtain genewise dispersion estimates, $\hat{\psi}_g$, and their generalized linear model methods to obtain estimates for the remaining gene-specific parameters, i.e. $\hat{\phi}_g$, $\hat{\alpha}_g$, and $\hat{\delta}_g$. Using $\hat{\theta}_g = (\hat{\phi}_g,\hat{\alpha}_g,\hat{\delta}_g,\hat{\psi}_g)$, we estimate hyperparameters for the location and scale parameters in the hierarchical model using a central method of moments approach, e.g. $\hat{\eta}_\phi = \sum_{g=1}^G \hat{\phi}_g/G$ and $\hat{\sigma}_\phi^2 = \sum_{g=1}^G (\hat{\phi}_g - \hat{\eta}_\phi)^2/(G-1)$ (and similarly for $\hat{\eta}_\psi$ and $\hat{\sigma}_\psi$) and $\hat{\eta}_\alpha = \sum_{g=1}^G \hat{\alpha}_g/G$ and $2\hat{\sigma}_\alpha^2 = \sum_{g=1}^G (\hat{\alpha}_g - \hat{\eta}_\alpha)^2/(G-1)$ (and similarly for $\hat{\eta}_\delta$ and $\hat{\sigma}_\delta$).

@@

To perform this Bayesian analysis, we used the statistical software \Stan{} which implements a variant of Markov chain Monte Carlo (MCMC) called Hamiltonion Monte Carlo (HMC) \citep{neal2011mcmc} to obtaining samples from the posterior in equation \eqref{e:condind}. We ran 4 simultaneous chains with random initial starting values for 1000 burnin (and tuning) iterations followed by another 1000 iterations thinned by 4 (to reduce storage space) for inference. We monitored convergence using the potential scale reduction factor and effective sample size (ESS) for $\phi_g$, $\alpha_g$, $\delta_g$, and $\psi_g$ \citep{Gelm:Rubi:infe:1992}. If the minimum ESS was less than 1000, we reran the chains with double the iterations for both burnin and inference. We continued this restarting and doubling until we obtained minimum ESS greater than 1000 for all parameters.

@@

\subsection{Implementation in \ShrinkBayes{}} \label{s:shrinkbayes}

@@

We used the methods described at the beginning of Section \ref{s:ebayes} to obtain normalization factors $\hat{\gamma}$ and gene-specific parameter estimates $\hat{\theta}_g$ for all genes using the \edgeR{} package \citep{robinson2010edgeR} from Bioconductor \citep{gentleman2004bioconductor} applied to the maize data on 27,888 genes with average count at least one and at most two zero read counts for each variety across the four replicates.  This analysis produced sample-specific normalization factors of $\hat{\gamma}=
(0.074, -0.059, -0.074, -0.014, -0.014, -0.124,  0.093,  0.063,  0.021, -0.037,  0.049,  0.021)$. The  gene-specific parameter estimates were treated as the true parameter values for the simulation study so that our simulated datasets mimicked the existing structure among the gene-variety means of the original maize data. If $\hat{\delta}_g<-|\hat{\alpha}_g$ ($\hat{\delta}_g >|\hat{\alpha}_g|$)  the gene was considered to have LPH (HPH).

Using these parameters and normalization factors, we simulated data according to the negative binomial model in \eqref{e:data} independently for each gene. For each simulation, we analyzed a subset of 25,000 genes selected randomly from genes with simulated counts at least one on average and with at most two zeros for each variety across replications.  We repeated this simulation process 10 times for each of 4, 8, and 16 replicates per variety, reusing normalization factors when necessary.  
% From Dan "Did we consider changing the number of zeros allowed when the number of reps is higher?  I like the idea of simplifying all this by just picking an average count threshold that his high enough to avoid problems.  I think that may be what others have done in the past."

@@

\edgeR{} can be used to test for differential expression between any two varieties based on the fit of a negative binomial log-linear model \citep{robinson2007moderated, robinson2010edgeR}. To construct a measure of heterosis, we computed the maximum likelihood estimates of the $\mu_{gv}$ parameters for all genes using \edgeR{}'s built-in Fisher scoring algorithm, and then used likelihood ratio tests to calculate two $p$-values for each gene: $p_{g1}$  for testing $H_{g01}:\mu_{g1}=\mu_{g3}$ and $p_{g2}$ for testing $H_{g02}:\mu_{g2}=\mu_{g3}$. Then, for each gene, we define a new $p$-value denoted as $p_g$ and set to $p_g=1$ if the estimate of $\mu_{g3}$ falls between the estimates of $\mu_{g1}$ and $\mu_{g2}$ and $p_g=\max\{p_{g1},p_{g2}\}/2$ otherwise.  (For all relevant significance thresholds $\omega$ near 0, it can be shown that rejecting the null hypothesis of no heterosis of expression for gene $g$ whenever $p_g\leq \omega$ results in a test that is asymptotically size $\omega$.)  We then use $1-p_g$ as a final measure positively associated with strength of evidence for gene expression heterosis.

{\tt baySeq} allows for a wider range of hypotheses for each gene, including $H^*_{g1}: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H^*_{g 2}: \mu_{g1}=\mu_{g2}$, $H^*_{g 3}: \mu_{g1}=\mu_{g3}$, $H^*_{g 4}: \mu_{g2}=\mu_{g3}$, and $H^*_{g 5}:$ all $\mu_g$'s are distinct. In a technique similar to our application of \edgeR{}, we used the posterior probabilities of these hypotheses to construct a measure of heterosis for each gene. We set this measure to zero if the maximum likelihood estimate, calculated using \edgeR{}, of $\mu_{g3}$ is between the maximum likelihood estimates of $\mu_{g1}$ and $\mu_{g2}$. Otherwise, the measure is the sum of the posterior probabilities of $H^*_{g 2}$ and $H^*_{g 5}$, the two hypotheses that allow for extreme heterosis.

@@

For each of the methods above, as well as the method described in Section \ref{s:gene_heterosis}, we sorted genes according to the computed measure of the strength of evidence for heterosis. From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to distinguish genes with heterosis, as defined in Section \ref{s:sim_data}, 
% From Dan, "why this citation?"
from those without heterosis \citep{landau2013dispersion}. A representative set of ROC curves is shown in Figure \ref{f:roc}. 
\begin{figure}[htbp]
\centerline{\includegraphics[width=\textwidth]{exampleROC0_1}}
\caption{Representative ROC curves for false positive rates below 0.1 for the approaches using \edgeR{} and \baySeq{} described in Section \ref{s:alternative}, the \ShrinkBayes{} approach described in Section \ref{s:shrinkbayes} and our eBayes approach described in Section \ref{s:ebayes}.}
\label{f:roc}
\end{figure}
For this simulation, we can see that the approaches based on the model in Section \ref{s:model}, i.e. eBayes and \ShrinkBayes{}, provide the best true positive rate for a given false positive rate. Also, as expected, as the sample size increases, our ability to distinguish genes with heterosis from genes without improves.

Figure \ref{f:auc} provides the area under the ROC curve below a false positive rate of 0.1 across the 10 simulations for each of the 3 different sample sizes. 
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{auc-facet-TRUE}}
\caption{Area under the ROC curves below a false positive rate of 0.1 for 3 different replicates per variety for the approaches using \edgeR{} and \baySeq{} described in Section \ref{s:alternative}, the \ShrinkBayes{} approach described in Section \ref{s:shrinkbayes} and our eBayes approach described in Section \ref{s:ebayes}. Each line is a different simulation while the blue box indicates mean AUC (plus or minus one standard error).}
\label{f:auc}
\end{figure}
The pattern appears to hold, with the eBayes and \ShrinkBayes{} methods outperforming the other two. With 4 replicates per variety, there does not appear to be much of a difference between \ShrinkBayes{} and our eBayes approach, but as the number of replicates increases our eBayes approach appears to improve relative to \ShrinkBayes{}.

@@

We used our method to analyze a maize data set \citep{paschold2012complementation} of \RNAseq{} gene expression in parental lines B73 and Mo17 and the hybrid genotype (B73$\times$Mo17). The data are available in the Supplemental Materials. Each variety had four biological replicates measured with Illumina methodology and equipment. Reads were mapped to the whole reference genome using the short reads aligner, NOVOALIGN. For more specifics, please see \cite{paschold2012complementation}.

As explained in Section~\ref{s:sim_data}, we analyzed a set of $G=27,888$ genes for $V=3$ varieties and $n_v=4$ replicates ($v=1,2,3$). We estimated the hyperparameters to be $\hat{\eta}_\psi = 3.44$, $\hat{\sigma}_\phi = 2.74$, $\hat{\eta}_\alpha = -0.05$, $\hat{\sigma}_\alpha=0.36$, $\hat{\eta}_\delta = 0.08$, $\hat{\sigma}_\delta=0.33$, $\hat{\eta}_\psi = -2.35$, and $\hat{\sigma}_\psi=0.60$.

With posteriors for all parameters, we can calculate empirical Bayes posterior probabilities for LPH and HPH. For each gene, the quantity of interest is likely to be the maximum of these two probabilities. For each gene with a high probability of either HPH or LPH, the magnitude of the effect is of interest. We calculate 
\begin{equation}
% \mbox{estimated effect size}_g = \left\{ 
% \begin{array}{ll}
% \hat{\mu}_{g3} - \min(\hat{\mu}_{g1},\hat{\mu}_{g2}) & \mbox{if } \hat{\mu}_{g3} < \min(\hat{\mu}_{g1},\hat{\mu}_{g2}) \\
% \hat{\mu}_{g3} - \max(\hat{\mu}_{g1},\hat{\mu}_{g2}) & \mbox{if } \hat{\mu}_{g3} > \max(\hat{\mu}_{g1},\hat{\mu}_{g2}) \\
% 0 & \mbox{otherwise},
% \end{array} 
% \right. 
\mbox{estimated effect size}_g \equiv \left\{ 
\begin{array}{ll}
\hat{\delta}_g - |\hat{\alpha}_g| & \mbox{if } \hat{\delta}_g > \phantom{-}|\alpha_g| \\
\hat{\delta}_g + |\hat{\alpha}_g| & \mbox{if } \hat{\delta}_g < -|\alpha_g| \\
0 & \mbox{otherwise},
\end{array}.
\right. 
\label{e:effect_size}
\end{equation}
Thus, the estimated effect size is the difference between hybrid mean and the nearest parent with negative values indicating LPH and positive values indicating HPH. If the hybrid mean is estimated to be between the parents, this effect is defined to be zero.

Figure \ref{f:volcano} provides a volcano plot, in this case a bivariate histogram, to visualize the maximum of the probabilities of LPH and HPH versus estimated effect size. 
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{volcano}}
\caption{A bivariate histogram of the maximum of the LPH and HPH probabilities versus estimated effect size defined in equation \ref{e:effect_size} for the B73 $\times$ Mo17 maize experiment.}
\label{f:volcano}
\end{figure}
The figure shows a peak at zero for probabilities below 0.5 with estimated effect sizes of zero. Above a probability of 0.5, we see a prototypical volcano pattern, with increased heterosis probability corresponding to larger estimated effect sizes and no estimated effect sizes near zero for genes with high heterosis probability. We also see asymmetry, with larger negative effect sizes than positive effect sizes due to genes with hybrid counts near zero and relatively high parental counts. Genes with high estimated posterior probabilities of heterosis and large estimated effect sizes are candidates for further investigation.

@@

With almost 30,000 genes in our data set, we can evaluate the distribution assumptions for the gene-specific parameters. Figure \ref{f:scatterplot} provides marginal and bivariate histograms for the gene-specific parameters.
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{estimates}}
\caption{Marginal and bivariate histograms for gene-specific parameter posterior meansfor the B73 $\times$ Mo17 maize experiment. For reference, the marginal histograms have a dashed line indicating the appropriate marginal probability density functions in the hierarchical model with the MAP estimates.}
\label{f:scatterplot}
\end{figure}

Geneticists speculate that gene expression heterosis is one possible explanation of phenotypic heterosis of traits, such as plant height or grain yield. Existing methods for identifying differential gene expression based on \RNAseq{} data are not directly applicable for detecting heterosis genes. \cite{ji2014estimation} introduced an empirical Bayes approach based on a hierarchical model for microarray data. We followed their approach, modified to allow for \RNAseq{} read counts as measures of transcript abundance. We developed an empirical Bayes approach based on obtaining estimates for hyperparameters followed by MCMC to estimate gene-specific parameters. The empirical Bayes posteriors can be used to estimate posterior probabilities of high and low parent heterosis. Through a simulation study, we demonstrated that this method outperformed alternative methods (that are not designed to detect heterosis), and performed comparably well with an implementation of our model in \ShrinkBayes{}, which estimates the full posterior via INLA. We then demonstrated the use of the methodology on a maize experiment in which phenotypic heterosis is well known.

% Estimates of heterosis to compare with Ji et. al.
In addition to providing estimates of heterosis probabilities and effect sizes, this real data analysis exposed violations of our normality assumptions, indicating room to improve the model and increase the power to detect gene expression heterosis. Some of our ongoing and future research will develop the details of such an extension.
